---
output:
  pdf_document: default
  html_document: default
---
```{r, results=FALSE}
library(readr)
library(astsa)
library(ggplot2)
library(forecast)
library(tidyverse)
library(lubridate)
library(dplyr)
library(car)
library(caret)
```

# ------------------------------------------
## Fitted ARIMA model by merging two datasets
# ------------------------------------------

```{r}
# Read and manipulate Cocoa Price Data

cocoa <- read.csv("Daily Prices_ICCO.csv", stringsAsFactors = FALSE)

cocoa$Date <- as.Date(cocoa$Date, format='%d/%m/%Y')
cocoa <- cocoa %>% arrange(Date)   # Ensure data is sorted by Date.

# Adjust the column names for convenience.
colnames(cocoa)[1] <- "DATE"
colnames(cocoa)[2] <- "DailyPrice"

cocoa$DailyPrice <- gsub(",", "", cocoa$DailyPrice) # Remove any non-numeric characters in Price column.
cocoa$DailyPrice <- as.numeric(cocoa$DailyPrice)    # Convert the results to numeric.

# Aggregate cocoa_data if there are multiple prices per day
cocoa_daily <- cocoa %>%
  group_by(DATE) %>%
  summarize(cocoa_price = mean(DailyPrice, na.rm = TRUE)) %>%
  ungroup()
```

```{r}
# Read and manipulate Weather and Precipitation Data

climate_data <- read.csv("Ghana_data.csv", stringsAsFactors = FALSE) %>%
  mutate(DATE = as.Date(DATE, format = "%Y-%m-%d")) %>%
  arrange(DATE)

climate_data$PRCP[climate_data$PRCP == ""] <- NA
climate_data <- climate_data %>%
  filter(!is.na(PRCP)) %>%
  filter(!is.na(TAVG))

climate_data$PRCP <- as.numeric(climate_data$PRCP)
climate_data$TMAX <- as.numeric(climate_data$TMAX)
climate_data$TMIN <- as.numeric(climate_data$TMIN)

# Aggregate climate_data if there are multiple stations per day
climate_daily <- climate_data %>%
  group_by(DATE) %>%
  summarize(
    prcp = mean(PRCP, na.rm = TRUE),
    tavg = mean(TAVG, na.rm = TRUE),
    tmax = mean(TMAX, na.rm = TRUE),
    tmin = mean(TMIN, na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r}
# Merge datasets
merged_data <- inner_join(cocoa_daily, climate_daily, by = "DATE")

# Split the dataset into training and testing datasets
n <- nrow(merged_data)
train_merged <- merged_data[1:(n - 5), ]
test_merged <- merged_data[(n - 4) : n, ]
```

```{r}
# Create time series objects
price_ts <- ts(merged_data$cocoa_price, frequency = 365)

# Calculate Variance Inflation Factor (VIF)
vif(lm(cocoa_price ~ prcp + tavg, data = merged_data))
```

We obtain the VIF value for average temperature and perception lower than 5, which indicates no significant multicollinearity.

```{r}
plot(price_ts)
acf2(price_ts)
```

The plot reveals non-stationarity with an upward trend and significant volatility, especially toward the end of the series, and the corresponding ACF plot indicates a slow decay of autocorrelations, confirming non-stationarity.

```{r}
ls = log(price_ts); dls = diff(ls)
plot.ts(cbind(price_ts,ls,dls), main="")
```

The log transformation is applied for variance-stabilizing, and we difference the log-transformed series to remove the trend.

```{r}
acf2(ls)
acf2(dls)
```

```{r}
# Fit Non-Seasonal ARIMA to the training data
train_ts <- ts(train_merged$cocoa_price, frequency = 365)

train_merged$Month <- factor(month(train_merged$DATE),levels = 1:12, labels = month.abb)
climate_reg <- model.matrix(~ Month - 1, data = train_merged)
climate_reg <- climate_reg[, -1]

arima_model_reg <- auto.arima(train_ts, seasonal = TRUE, xreg = climate_reg)
summary(arima_model_reg)
```

```{r}
sarima(train_ts, 4, 1, 1)
```


```{r}
#Further Predictions and Comparison with Test Data
test_merged$Month <- factor(month(test_merged$DATE), levels = 1:12, labels = month.abb)
forecast_reg <- model.matrix(~ Month - 1, data = test_merged)[, -1]
forecast_arima_reg <- forecast(arima_model_reg, h = 5, xreg = forecast_reg)
forecast_arima_reg
forecast_arima <- forecast(arima_111, h = 5)
forecast_arima

test_ts <- ts(test_merged$cocoa_price, frequency = 365)
test_ts

forecast_df <- data.frame(
  Date = test_merged$DATE,
  Actual = test_merged$cocoa_price,
  ARIMA_Reg_Forecast = forecast_arima_reg$mean,
  ARIMA_Forecast = forecast_arima$mean)

# Plot actual vs predicted values
ggplot(forecast_df, aes(x = Date)) +
  geom_line(aes(y = Actual, color = "Actual Price"), linewidth = 1) +
  geom_line(aes(y = ARIMA_Reg_Forecast, color = "ARIMA Reg Forecast"), linetype = "dotted", linewidth = 1) +
  geom_line(aes(y = ARIMA_Forecast, color = "ARIMA Forecast"), linetype = "dotted", linewidth = 1) +
  labs(title = "Model Prediction vs. Actual Cocoa Prices (Test Data)",
       x = "Date",
       y = "Cocoa Price",
       color = "Legend") +
  theme_minimal()
```
